\addchap{Acknowledgments}
%\begin{refsection}
%
%%content goes here
%
%%\printbibliography[heading=subbibliography]
%\end{refsection}

This grammar has been written with the help of colleagues, friends and family. I wish to thank in particular:

\begin{itemize}
	\item Tshendzin \zh{陈珍} and  Dpalcan \zh{柏尔青}, who taught me Japhug.
	\item Andzin, Ayang/Kunbzang Mtsho,   Tshering Skyid, for telling me stories in Japhug.
	\item Norbu \zh{罗尔武} and Pengfeng \zh{彭锋}, for their help and support during my stays in Mbarkham.
	\item Jackson T.-S. Sun \zh{孙天心} and Lin Youjing \zh{林幼菁}, from whom I have learnt a lot, especially during our time together doing fieldwork in Mbarkham and Chengdu.	
	\item Laurent Sagart, for his invaluable help for preparing my first fieldwork trip, and his guidance in the early part of my career.
	\item 	Yang Dongfang \zh{杨东方} and Chang Yueh-chin \zh{张月琴}, for their help during my first stay in Mbarkham in 2002, and for introducing me to Tshendzin.
	\item My former PhD students Lai Yunfan \zh{赖云帆}, Gong Xun \zh{龚勋} and Zhang Shuya \zh{章舒娅}, for their challenging questions and comments, and for helping recording some of the texts in the corpus.
	\item Alexis Michaud, Boyd Michailovksy and Séverine Guillaume, without whom I would not have made my Japhug texts available on the Pangloss archive.
	\item Céline Buret, Rémi Bonnet, Benjamin Galliot and Alexander Delaporte, who helped producing several versions of the Japhug-French-Chinese dictionary.
	\item Martin Haspelmath, for accepting this manuscript in the series `Studies in Diversity Linguistics' and for organizing the evaluation process.
	\item Thomas Pellard and Sebastian Nordhoff, for their guidance with \LaTeX.
	\item Anton Antonov, who corrected no less than six chapters (\ref{chap:nominal.morphology}, \ref{chap:verb.template}, \ref{chap:nonconcatenative}, \ref{chap:negation}, \ref{chap:indexation}, \ref{chap:orientation.AM}, \ref{chap:other.derivations})
	\item the LSP reviewers: Jeroen van de Weijer (chapters \ref{chap:phono} and \ref{sec:clusters.redp}), Tom Bossuyt (\ref{chap:pronouns}), X (\ref{chap:numerals}), X (\ref{chap:postpositions.relators}), X (\ref{chap:noun.phrase}), X (\ref{chap:ideophones.sfp}), Ye Jingting \zh{叶婧婷} (\ref{chap:non-finite}), X (\ref{chap:valency.increasing.derivation}), Kasia Janic (\ref{chap:valency.decreasing.derivation}), Aimée Lahaussois (\ref{chap:denominal}), X (\ref{chap:tame}), X (\ref{sec:basic.syntax}), Valérie Guérin (\ref{chap:relatives}), X (\ref{chap:complement.clauses}) and Yvonne Treis (\ref{chap:degree}), as well those who prefer to remain anonymous.
	\item Theo Yeh \zh{葉庭英}, Roland Pooth and Marc Miyake, who provided useful corrections on the draft.
	\item Archi and Mira, for being there.
\end{itemize}
